Supporting Information : Asymmetric tail-dependence in extreme climatic events affects population-extinction risk

### 2019-08-30 MainText.Rmd
discussion edited

<!--how common are the extreme events?-->
For past few decades, extreme climatic events (ECE) appeared more common than 
usual due to anthropogenic climate change, e.g. temperature or precipitation extremes,
heat-waves, wild-fire etc. [@ummenhofer2017extreme]. Frequent extreme events 
can shift temperature or precipitation-mean, 
change variability of the distribution, as well as change its 
symmetry or skewness [@hansen2012perception]. Extreme climatic events can pose severe
impacts at the individual, population, community and ecosystem level [@bragazza2008climatic; 
@gutschick2003extreme; @jiguet2011community; @mckechnie2009climate;
@hoover2014resistance]. Therefore, 
understanding the long-term impacts of changes 
in ECEs has been a longstanding interest in ecological research 
[@katz2005statistics; @smith2011ecological; @diez2012will; @moritz2013future; @bailey2016tackling].

@@

We simulated our metapopulation models with noises $(\epsilon_t^i)$ that were 
independent through time with same spatial correlations between patches (to within sampling error)
for every simulation, but moderately (or extremely) left- and right-tail dependent between patches in
different simulations. For moderately left-tail dependent noise we drew points from a $n$ dimensional 
Clayton copula with same Kendall correlation ($= 0.6$) between any two out of $n$ patches and made 
the marginals normal with zero mean and standard deviation $\sigma$ 
(see the specified value $\sigma$ for each model in 3$^{rd}$ 
column of Table \ref{tab_model_summary}). For moderately 
right-tail dependent noise we flipped ($180^{\circ}$ rotation) the points drawn 
from Clayton copula so that all the statistical properties remained 
same except for the direction of tail-dependence. We generated
extreme left- or right-tail dependent noises via algorithm stated in 
Appendix \ref{SM-extr_tail_dep}. Our models (Table \ref{tab_model_summary}) 
computed extinction risk for $n_t = 25$ years, $n = 5$ patches
and $n_s = 10000$ simulations with both moderate and extreme tail dependence (left-, right-tail).

@@

<!--summary results:-->
Our results show the potential use of "asymmetric" copula in estimating the impact of 
extreme ecological events on population-extinction risk. We show that through left and right tail 
dependence in vital rates (respectively, spatially correlated environmental noises), years of extreme events 
("catastrophs" and "bonanzas") affect extinction risk of a single-patch age-structured population
(respectively, density dependent metapopulations) differently. Though left tail dependence is 
expected to show higher extinction risk than the one for right tail dependence for age-structured 
and density dependent metapopulation models, it is not always true 
when metapopulation models with dispersal show over-compensatory dynamics. In the over-compensatory regime,
right tail dependent noises result in higher extinction risk than the left tail dependent one. We show 
that in absence of dispersal, the above-said switching in extinction risk measure occurs near the 
critical parameter value where the deterministic model shows transition from 
under-compensatory to over-compensatory regime.

<!--Tuljapurkar approximation : linear version of age-structured model in stochastic environments-->
We found left tail dependence in vital rates (induced by ECEs) caused higher extinction risk than 
the right tail dependent ones, but the difference was not remarkable (Fig. \ref{SM-fig_extrisk_agestr_mtd}) 
unless the tail dependence was at its extreme (Fig. \ref{fig_extrisk_agestr_etd}). This finding could be explained 
by the approximation @tuljapurkar1990lecture made for the stochastic log(growth rate) in an iid environment.
The approximation considered that the growth rate depends on the most dominant eigen value of the mean projection matrix 
and the temporal variance of the log(population growth rate) caused by environmental stochasticity, which was based
on the linear covariance measure between any two matrix elements, not on their tail-dependence properties. Therefore,
according to Tuljapurkar's approximation, population growth rate in an iid environment 
remained same for our model (Eq. \ref{eq_age_str}) with left-tail or right-tail dependent noises.

<!--future study: worthy for PVA, IUCN red list
1. Identify which life stage or demographic process as management target?
2. Old study (Doak1994, Ferguson1995) showed including correlation btw vital rates for PVA makes a difference, 
   what if we include TD through copula? Multisite age str. model with dispersal, directional change
   in spatio-temporal variation for vital rates.
3. Tdep. with -ve correlation between vital rates?
4. extinction risk how propagated through a food web?
5. ref: see McCarthy2000: Most previous models of metapopulation viability have either ignored
correlated extinction of patches, or have included uniform correlations, with the correlation between nearby patches being the same as the correlation between distant
patches (Hanski 1994; Tilman et al. 1994; Lindenmayer et al. 1995; McCarthy
1996). 
6. see this: http://dx.doi.org/10.1098/rspb.2015.0173
-->
Although we demonstrated that tail dependence between vital rates of an age structured model can
influence extinction risk, substantial work remains to determine the importance of this effect in the 
following scenarios. First, we demonstrated the effect of tail-dependence on extinction risk 
with a very simplistic age structured (two age stages) model where we considered the total semelperous 
population would go extinct if there would be only one female adult and one egg. But in reality population 
viability analysis (PVA) deals with a little bit higher threshold with several age, stage or size class and a variety 
of populations including semelperous and iteroperous [@morris2002quantitative]. Future studies should include 
tail-dependence effect on extinction risk 
for a multiple age-, size- and stage- structured model and estimate extinction risk 
for each age, size and stage class separately 
so that we can identify (via multisite-PVA) which life-stage or demographic process 
to target for population management. Earlier study by @doak1994modeling showed PVA models including correlation 
between different demographic rates indicated marked difference in predicting confidence interval for 
future population size than the model which did not include correlation. But when 
such correlation cannot be estimated directly, more generally where the
forms of statistical dependence among the model's parameters are unknown, making incorrect
assumptions about correlations and dependencies may lead to severe over-estimation 
or under-estimation of extinction risks [@ferson1995correlations]. We believe modeling 
dependencies with copula could help in this case. Second, we did not consider multiple sites of 
age-structured metapopulation which could be linked via dispersal. So, possible future investigation 
would be to explore the viability for the metapopulations if vital rates (e.g. fecundity or 
survival of different stage class) across sites showed specific tail dependence 
with spatially correlated extreme climatic events. Third, we considered tail dependence 
between vital rates when they were positively correlated, but what would be the extinction risk 
and viability of a population if demographic rates were negatively correlated with asymmetric tail dependence. 
This scenario could be considered for those extreme events that kill adults and allow more successful
recruitment of seeds or other propagules. Fourth, following the aforementioned idea, it would be worthy 
to model more complex prey-predator model where asymmetric tail dependent copula could be used to mimic
interspecific-dependence. An earlier study [@carnicer2011widespread] showed extreme events
(i.e., heat waves and droughts) increased mortality and defoliation in southern European forest and its 
dependent herbivorous insects and fungi through trophic linkage. ECEs can even truncate 
multiyear insect outbreak dynamics [@esper20061200]. We believe copula (may be a multidimensional one) has 
a lot more potential to estimate risk of cascading extinction in model food webs.

<!--match-mismatch hypothesis leads to extinction? our phenology paper in prep.-->

>>>>

<!--how common are the extreme events?-->
For past few decades, extreme climatic events (ECEs) appeared more common than 
usual due to anthropogenic climate change, e.g. temperature or precipitation extremes,
heat-waves, wild-fire etc. [@ummenhofer2017extreme]. Frequent extreme events 
can shift temperature or precipitation-mean, 
change variability of the distribution, as well as change its 
symmetry or skewness [@hansen2012perception]. Extreme climatic events can pose severe
impacts at the individual, population, community and ecosystem level [@bragazza2008climatic; 
@gutschick2003extreme; @jiguet2011community; @mckechnie2009climate;
@hoover2014resistance]. Therefore, 
understanding the long-term impacts of changes 
in ECEs has been a longstanding interest in ecological research 
[@katz2005statistics; @smith2011ecological; @diez2012will; @moritz2013future; @bailey2016tackling].

@@

We simulated our metapopulation models with noises $(\epsilon_t^i)$ that were 
independent through time with same spatial correlations between patches (to within sampling error)
for every simulation, but moderately (or extremely) left- and right-tail dependent between patches in
different simulations. For moderately left-tail dependent noise we drew points from a $n$ dimensional 
Clayton copula with same Kendall correlation ($= 0.6$) between any two out of $n$ patches and made 
the marginals normal with zero mean and standard deviation $\sigma$ 
(see the specified value $\sigma$ for each model in 3$^{rd}$ 
column of Table \ref{tab_model_summary}). For moderately 
right-tail dependent noise we flipped ($180^{\circ}$ rotation) the points drawn 
from Clayton copula so that all the statistical properties remained 
same except for the direction of tail-dependence. We generated
extreme left- or right-tail dependent noises via algorithm stated in 
Appendix \ref{SM-extr_tail_dep}. Our models (Table \ref{tab_model_summary}) 
computed extinction risk after $25$ years for $10000$ simulations at $n = 5$ patches
with spatially correlated noises having either moderate or extreme tail 
dependence in left and right-tail.

@@

<!--summary results:-->
We showed the potential use of "asymmetric" tail dependence in modeling the impact of 
extreme ecological events on population-extinction risk. Our results demonstrated left and right-tail 
dependence in vital rates (or in spatially correlated environmental noises) affected extinction risk differently
in years of extreme events ("catastrophs" and "bonanzas"). Though we could expect left tail dependence 
to show higher extinction risk than the one for right tail dependence with age-structured 
and density dependent metapopulation models, it was not always true. In the over-compensatory regime,
right tail dependent noise caused higher extinction risk than the left tail dependent noise. We showed 
that in absence of dispersal the above-said switching took place near the critical parameter value where 
the deterministic model showed transition from under-compensatory to over-compensatory regime.

<!--Tuljapurkar approximation : linear version of age-structured model in stochastic environments-->
We found left tail dependence in vital rates (induced by ECEs) caused higher extinction risk than 
the right tail dependent ones, but the difference was not remarkable (Fig. \ref{SM-fig_extrisk_agestr_mtd}) 
unless the tail dependence was at its extreme (Fig. \ref{fig_extrisk_agestr_etd}). This finding could be explained 
by the approximation @tuljapurkar1990lecture made for the stochastic log(growth rate) in an iid environment.
The approximation considered growth rate being dependent on the most dominant eigen value of the mean projection matrix 
and the temporal variance of the log(population growth rate) caused by environmental stochasticity - which was based
on the linear covariance measure between any two matrix elements, not on their tail-dependence properties. Therefore,
following @tuljapurkar1990lecture, population growth rate in an iid environment could be approximated
as same for our model (Eq. \ref{eq_age_str}) with left-tail or right-tail dependence.

<!--future study: worthy for PVA, IUCN red list:
1. Identify which life stage or demographic process as management target?
2. Old study (Doak1994, Ferguson1995) showed including correlation btw vital rates for PVA makes a difference, 
   what if we include TD through copula? Multisite age str. model with dispersal, directional change
   in spatio-temporal variation for vital rates.
3. Tdep. with -ve correlation between vital rates?
4. extinction risk how propagated through a food web?
5. Not sure : see McCarthy2000: Most previous models of metapopulation viability have either ignored
correlated extinction of patches, or have included uniform correlations, with the correlation between 
nearby patches being the same as the correlation between distant
patches (Hanski 1994; Tilman et al. 1994; Lindenmayer et al. 1995; McCarthy
1996). 
-->
Although we demonstrated that tail dependence between vital rates of an age structured model can
influence extinction risk, substantial work remains to determine the importance of this effect in the 
following scenarios. First, we demonstrated the effect of tail-dependence on extinction risk 
with a basic age structured (two age stages) model where we considered the total semelperous 
population would go extinct if there would be only one female adult and one egg. But in reality population 
viability analysis (PVA) deals with a little bit higher threshold with several age, stage or size class and a variety 
of populations including semelperous and iteroperous [@morris2002quantitative]. Future studies should include 
tail-dependence effect on extinction risk 
for a multiple age-, size- and stage- structured model and estimate extinction risk 
for each age, size and stage class separately 
so that we can identify (via multisite-PVA) which life-stage or demographic process 
to target for population management. Earlier study by @doak1994modeling showed PVA models with correlation 
between different demographic rates indicated marked difference in predicting confidence interval for 
future population size compared with the model which did not include correlation. But when 
such correlation cannot be estimated directly, more generally where the
forms of statistical dependence among the model's parameters are unknown, making incorrect
assumptions about correlations and dependencies may lead to severe over-estimation 
or under-estimation of extinction risks [@ferson1995correlations]. We believe modeling 
dependencies with copula could help in this case. Second, we did not consider multiple sites of 
age-structured metapopulation which could be linked via dispersal. So, possible future investigation 
would be to explore the viability for the metapopulations if vital rates (e.g. fecundity or 
survival of different stage class) across sites showed specific tail dependence 
with spatially correlated extreme climatic events. Third, we considered tail dependence 
between positively correlated vital rates, but what would be the extinction risk 
and viability of a population if demographic rates were negatively correlated with asymmetric tail dependence. 
We could consider this scenario to model extreme events that kill adults and allow more successful
recruitment of seeds or other propagules. Fourth, following the aforementioned idea, it would be worthy 
to model more complex prey-predator model where asymmetric tail dependent copula could be used to mimic
interspecific-dependence. An earlier study [@carnicer2011widespread] showed extreme events
(i.e., heat waves and droughts) increased mortality and defoliation in southern European forest and its 
dependent herbivorous insects and fungi through trophic linkage. <!--ECEs can even truncate 
multiyear insect outbreak dynamics [@esper20061200].--> We believe copula (may be a multidimensional one) has 
a lot more potential to estimate risk of cascading extinction in model food webs. <!--should I mention about 
match-mismatch hypothesis leads to extinction? our phenology paper in prep.--> Fifth, we considered 
our metapopulation patches were equispaced on a linear chain with equal global dispersal 
rate between patches. But in realistic network models, as inter-patch distances 
can be different, the dispersal rate should depend on distance: dispersal is possible at higher 
rates between nearby patches than the patches which are far away. It is worthy to study 
metapopulation viability with non-uniform spatial correlation (i.e. asymmetric tail dependence) in a 
network with different clusters having various dispersal rates.

### 2019-08-30 SuppMat.Rmd
discussion edited

For extreme tail-dependence algorithm, a $N \times n$ matrix, M was generated of independent 
standard-normal random draws (in general). <!--where $N = numsteps(=25) \times numsims(=10000)$ : only for 2d case-->.
Then, each row of M was identified for a first set of modifications
randomly and independently with a 50\% chance; for those rows identified, every
element of the row was replaced with the absolute
value of the first element of the row (including the first element itself).
For other rows, every element in the row was replaced by the negative of 
its own absolute value. This produces right-tail dependent
matrix, normally distributed for each column. If left-tail 
dependent noise is desired, take the negative of the entire matrix.
Our [code](https://github.com/sghosh89/ERC/blob/master/ExtremeTailDep.R) for
generating extreme tail dependence matrix is flexible enough, so that 
matrix M can be generated initially with independent random draws from a normal 
distribution with a specified mean and standard deviation.

>>>>

For extreme tail-dependence algorithm, a $N \times n$ matrix, M was generated of independent 
standard-normal random draws (in general). <!--where $N = numsteps(=25) \times numsims(=10000)$ : only for 2d case-->
Then, each row of M was identified for a first set of modifications
randomly and independently with a 50\% chance; for those rows identified, every
element of the row was replaced with the absolute
value of the first element of the row (including the first element itself).
For other rows, every element in the row was replaced by the negative of 
its own absolute value. This produces right-tail dependent
matrix, normally distributed for each column. If left-tail 
dependent noise is desired, take the negative of the entire matrix.
Our [code](https://github.com/sghosh89/ERC/blob/master/ExtremeTailDep.R) for
generating extreme tail dependence matrix is flexible enough, so that 
matrix M can be generated initially with independent random draws from a normal 
distribution with a specified mean and standard deviation.
